Fail Faster

### 2025-03-19 paper/paper.tex
eval pass

The runtime of these generators over a range of sizes is shown in \figref{fig:time}. Staging provides only a negligible speedup for type-derived and insertion-based BST strategies, whose overhead is dominated by random sampling. In this case, performance improvement is primarily attributable the \csplitmix \rand.
In contrast, STLC strategies benefit more from staging, as they involve fewer sampling operations but rely heavily on generator combinators. The same holds for boolean lists, as sampling booleans is inexpensive relative to integers. Our single-pass BST strategy falls somewhere in between: it both samples frequently and makes extensive use of generator combinators. \tr{Goal is to replace this paragraph with the ref-counting chart---I think it will be WAY less boring.}

@@

As shown in \figref{fig:scala}, \scalaname achieves an even greater performance gain over ScalaCheck than \name does over \bq. This discrepancy arises from ScalaCheck's representation of generators as functions of type \texttt{size -> seed -> Option[A]}. Each generator combinator must construct and then pattern-match on these \texttt{Option} values, introducing significant boxing and unboxing overhead at each step.
Staging allows us to eliminate this overhead, resulting in substantial speedups.

These results illustrate that staging is a powerful \textit{and} general optimization technique, capable of delivering performance gains in strict functional languages beyond OCaml.
\begin{figure}[h]
  % \includegraphics[scale=0.33]{allocs.png}
  \includegraphics[scale=0.45]{time.png}
  \caption{Time to generate a value using each strategy, averaged over 10,000 trials. BQ is \bq; Staged SR is \name using \bq's \texttt{splittable\_random} RNG; Staged CSR is \name using \secref{subsection:faster-rng}'s \csplitmix. Performance benefits scale with generation size; lower is better.}
  \label{fig:time}
\end{figure}

@@

The average of individual-task speedups for each strategy and benchmark, across all seeds, is shown in \figref{fig:etna}. Trends in bug-finding ability closely reproduce trends in performance from \secref{subsection:benchmarking}. The distribution of individual-task speedups, shown in \figref{fig:swarm}, tell a more interesting story: benchmarks

\begin{figure}[h]
  \includegraphics[scale=0.333]{speedup.png}
  \caption{Geometric average of all speedups for each strategy and benchmark, showing that staging leads to better bug-finding ability across the board.}
  \label{fig:etna}
\end{figure}

\begin{figure}[h]
  \includegraphics[scale=0.41]{swarm.png}
  \caption{Distribution of individual-task speedups across strategies and benchmarks. Extreme outliers correspond to large speedups.}
  \label{fig:swarm}
\end{figure}

>>>>

The runtime of these generators over a range of sizes is shown in \figref{fig:time}. Staging provides only a negligible speedup for type-derived and insertion-based BST strategies, whose overhead is dominated by random sampling. In this case, performance improvement is primarily attributable the \csplitmix \rand.
In contrast, STLC strategies benefit more from staging, as they involve fewer sampling operations but rely heavily on generator combinators. The same holds for boolean lists, as sampling booleans is inexpensive relative to integers. Our single-pass BST strategy falls somewhere in between: it both samples frequently and makes extensive use of generator combinators. \tr{Goal is to replace this paragraph with the ref-counting chart---I think it will be WAY less boring. But I need to do 3.7 first.}

@@

As shown in \figref{fig:scala}, \scalaname achieves a greater performance gain over ScalaCheck than \name does over \bq. This discrepancy arises from ScalaCheck's representation of generators as functions of type \texttt{size -> seed -> Option[A]}. Each generator combinator must construct and then pattern-match on these \texttt{Option} values, introducing significant boxing and unboxing overhead at each step.
Staging allows us to eliminate this overhead, resulting in substantial speedups.

These results illustrate that staging is a powerful \textit{and} general optimization technique, capable of delivering performance gains in strict functional languages beyond OCaml.
\begin{figure}[h]
  % \includegraphics[scale=0.33]{allocs.png}
  \includegraphics[scale=0.45]{time.png}
  \caption{Time to generate a value using each strategy, averaged over 10,000 trials. Performance benefits scale with generation size; lower is better. BQ is \bq; Staged SR is \name using \bq's \texttt{splittable\_random} RNG; Staged CSR is \name using \secref{subsection:faster-rng}'s \csplitmix.}
  \label{fig:time}
\end{figure}

@@

The average individual-task speedups for each strategy and benchmark are shown in 
\figref{fig:etna}, with trends in bug-finding ability closely mirroring those in performance 
from \secref{subsection:benchmarking}. The distribution of speedups in \figref{fig:swarm} reveals 
substantial variability, with most tasks clustering near the median and a long tail of outliers 
achieving much larger gains. STLC shows a more bimodal distribution of speedups
than the other benchmarks, which we attribute to the heterogenous difficulty of its tasks:  
some tasks regularly hit the 60-second timeout, while others finish in a fraction of a second. We find that tasks of intermediate difficulty 
tend to achieve larger speedups, whereas very fast or slow tasks cluster near the mean.
Overall, 
these results give strong evidence that our technique consistently benefits bug-finding ability, sometimes drastically.
\begin{figure}[h]
  \includegraphics[scale=0.333]{speedup.png}
  \caption{Geometric average of all speedups for each strategy and benchmark, showing that staging leads to better bug-finding ability across the board.}
  \label{fig:etna}
\end{figure}

\begin{figure}[h]
  \includegraphics[scale=0.41]{swarm.png}
  \caption{Swarm distribution of individual-task speedups across strategies and benchmarks (each black dot is a task). Plots with extremely low variance have no swarm overlay; in plots with high variance, outliers correspond to large speedups. \tr{Make mean higher-contrast.}}
  \label{fig:swarm}
\end{figure}
